---
title: "Data Exploration: Columbia River Adult Salmon Returns"
author: "[Jim Tyhurst](https://www.jimtyhurst.com/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
numbersections: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Configuration
```{r}
library(salmonstats)
library(glue)
library(tidyverse)
```

## Data preparation

Daily counts were obtained by manually selecting search parameters from the [Fish Passage Center](http://www.fpc.org/)'s [Adult Data Query](http://www.fpc.org/web/apps/adultsalmon/Q_adultcounts_dataquery.php) page:

* _passage reporting site_ : This exploration uses observations from the [Bonneville Dam](https://www.nwp.usace.army.mil/bonneville/), which is abbreviated as 'BON' in the 'Dam' column of the data.
* _species_ : I always left the default value 'All Species'.
* _start date_ : I always selected 'January 1' of the selected year.
* _end date_ : I always selected 'December 31' of the selected year.

After downloading each file of data, the final two lines of each file needed to be deleted, because they were annotations that were _not_ part of the data to be analyzed. Then I changed the name of the files to match this pattern:

> fpc-salmon-{passage_reporting_site_abbreviation}-{year}0101-{year}1231.csv

For example, the following files are for data from Bonneville Dam (BON) from years 2000 and 2015, respectively:

> fpc_salmon-BON-20000101-20001231.csv  
> fpc_salmon-BON-20150101-20001231.csv

Metadata is available from the [Fish Passage Center Adult Metadata](http://www.fpc.org/documents/metadata/FPC_Adult_Metadata.html) web page.

```{r warning='hide'}
# Replace the path components with the location of your data.
data_directory <- file.path("~", "data", "salmon", "ColumbiaRiver", "FishPassageCenter", "BonnevilleDam")

bon2000 <- get_fpc_data(2000, "BON", data_directory)
bon2015 <- get_fpc_data(2015, "BON", data_directory)
```

## Plots for Bonneville Dam for years 2000 and 2015

```{r fig.height = 6, fig.width = 8, fig.align = "center"}
plot_single_fpc_file(bon2000, 2000, "Bonneville Dam")
```


```{r fig.height = 6, fig.width = 8, fig.align = "center"}
plot_single_fpc_file(bon2015, 2015, "Bonneville Dam")
```

## References

Source code: https://github.com/jimtyhurst/salmonstats

Data source: http://www.fpc.org/web/apps/adultsalmon/Q_adultcounts_dataquery.php

Metadata: http://www.fpc.org/documents/metadata/FPC_Adult_Metadata.html
