---
title: "Data Exploration: Columbia River Adult Salmon Returns"
author: "[Jim Tyhurst](https://www.jimtyhurst.com/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
numbersections: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Configuration
```{r}
library(salmonstats)
library(glue)
library(tidyverse)
```

## Data preparation

Daily counts were obtained by manually selecting search parameters from the [Fish Passage Center](http://www.fpc.org/)'s [Adult Data Query](http://www.fpc.org/web/apps/adultsalmon/Q_adultcounts_dataquery.php) page:

* _passage reporting site_ : This exploration uses observations from the [Bonneville Dam](https://www.nwp.usace.army.mil/bonneville/), which is abbreviated as 'BON' in the 'Dam' column of the data.
* _species_ : I always left the default value 'All Species'.
* _start date_ : I always selected 'January 1' of the selected year.
* _end date_ : I always selected 'December 31' of the selected year.

After downloading each file of data, the final two lines of each file needed to be deleted, because they were annotations that were _not_ part of the data to be analyzed. Then I changed the name of the files to match this pattern:

> fpc-salmon-{passage_reporting_site_abbreviation}-{year}0101-{year}1231.csv

For example, the following files are for data from Bonneville Dam (BON) from years 2000 and 2015, respectively:

> fpc_salmon-BON-20000101-20001231.csv  
> fpc_salmon-BON-20150101-20001231.csv

Metadata is available from the [Fish Passage Center Adult Metadata](http://www.fpc.org/documents/metadata/FPC_Adult_Metadata.html) web page.

```{r results='hide'}
# Replace the path components with the location of your data.
data_directory <- file.path("..", "inst", "extdata", "ColumbiaRiver-FishPassageCenter", "BonnevilleDam")

bon2000 <- get_fpc_data(2000, "BON", data_directory)
bon2015 <- get_fpc_data(2015, "BON", data_directory)
```

## Plots for Bonneville Dam for years 2000 and 2015

```{r fig.height = 6, fig.width = 8, fig.align = "center"}
plot_single_fpc_file(bon2000, 2000, "Bonneville Dam")
```


```{r fig.height = 6, fig.width = 8, fig.align = "center"}
plot_single_fpc_file(bon2015, 2015, "Bonneville Dam")
```

I am surprised that the numbers for 2015 are so much bigger than for 2000.

## Plots for Bonneville Dam from 1990 to 2018

Each year of data is in one CSV file in a subdirectory of `inst/extdata` with a separate directory for each passage reporting site. Let's look at data for steelhead salmon first:

```{r}
first_year <- 1990
last_year <- 2018
passage_reporting_site_name <- "Bonneville Dam"
passage_reporting_site_code <- "BON"

# Replace the path components with the location of your data.
data_directory <- file.path("..", "inst", "extdata", "ColumbiaRiver-FishPassageCenter", "BonnevilleDam")

steelhead <- first_year:last_year %>% 
  purrr::map(function(year) {get_fpc_data(year, passage_reporting_site_code, data_directory)}) %>% 
  purrr::reduce(function(x, y) {dplyr::bind_rows(x, y)}) %>% 
  dplyr::select(Dam, Year, DayOfYear, Steelhead)
dim(steelhead)
```

We plot a few years to get a feel for the distribution:

```{r fig.height = 6, fig.width = 8, fig.align = "center"}
steelhead %>% 
  filter(Year %in% c(2000, 2005, 2010, 2015)) %>% 
  ggplot(aes(DayOfYear)) +
  geom_line(aes(y = Steelhead, color = factor(Year))) +
  ggtitle(
    glue("{passage_reporting_site}: Daily Count of Adult {species} Returns",
         passage_reporting_site = passage_reporting_site_name,
         species = "Steelhead")) +
  scale_x_continuous(name = "Day of Year", breaks = seq(0, 350, by = 50)) +
  scale_y_continuous(name = "Daily Count", limits = c(0, 10000), breaks = seq(0, 10000, by = 2000))
```

I notice that 2010 and 2015 have higher peaks than 2000 and 2005, but I thought that numbers were descreasing. Let's look more closely at the yearly trend. In the next plot, we compare total quantity for the whole year.

```{r fig.height = 6, fig.width = 8, fig.align = "center"}
steelhead %>% 
  dplyr::group_by(Year) %>% 
  dplyr::mutate(Quantity = sum(Steelhead)) %>%
  ggplot(aes(Year, Quantity)) +
  geom_bar(stat = "identity") +
  ggtitle(
    glue("{passage_reporting_site}: Yearly Total of Adult {species} Returns",
         passage_reporting_site = passage_reporting_site_name,
         species = "Steelhead")) +
  scale_x_continuous(name = "Year", breaks = seq(1990, 2015, by = 5)) +
  scale_y_continuous(name = "Yearly Total", labels = function(x){format(x, scientific = FALSE, big.mark = ",")})
```

## References

Source code: https://github.com/jimtyhurst/salmonstats

Data source: http://www.fpc.org/web/apps/adultsalmon/Q_adultcounts_dataquery.php

Metadata: http://www.fpc.org/documents/metadata/FPC_Adult_Metadata.html
