---
title: "NuSEDS Adult Salmon Escapement Data"
author: "[Jim Tyhurst](https://www.jimtyhurst.com/)"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
numbersections: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

&#x1F53B;&nbsp;_Work in Process_ ...&nbsp;&#x1F53A;

Source code at: [nuseds-adult-salmon-escapement-data.Rmd](../vignettes/nuseds-adult-salmon-escapement-data.Rmd)

## Configuration

```{r}
library(salmonstats)
library(descriptr)
library(tidyverse)
```

## New Salmon Escapement Database System (NuSEDS)

[NuSEDS](https://open.canada.ca/en/suggested-datasets/new-salmon-escapement-database-nuseds) is the Pacific Region’s centralized Oracle database that holds adult salmon escapement data. About 10,000 salmon spawning sites in DFO Pacific Region have been observed 0 to 10 times per year for 6 species for nearly 100 years (poorly before 1948). Data is aggregated within year; the raw data is largely on paper.

* zip file containing NUSEDS_20180416.csv  
    http://www.pac.dfo-mpo.gc.ca/od-ds/science/sed-des/NUSEDS_20180416.zip
* meta data  
    https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6

```{r}
ns18 <- get_nused_data()
dim(ns18)
```

```{r}
descriptr::ds_screener(ns18)
```

## MAX_ESTIMATE
The maximum estimated number taken from: NATURAL_ADULT_SPAWNERS, NATURAL_JACK_SPAWNERS, NATURAL_SPAWNERS_TOTAL, ADULT_BROODSTOCK_REMOVALS, JACK_BROODSTOCK_REMOVALS,  TOTAL_BROODSTOCK_REMOVALS, OTHER_REMOVALS, TOTAL_RETURN_TO_RIVER, UNSPECIFIED_RETURNS.

```{r, echo=FALSE}
plot(ns18$MAX_ESTIMATE)
```

## References

Source code: [vignettes](../vignettes/)

NUSEDS_20180416.csv  
    http://www.pac.dfo-mpo.gc.ca/od-ds/science/sed-des/NUSEDS_20180416.zip

NuSEDS meta data  
    https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6
